Statistical analysis of periodic data in neuroscience

### 2021-03-28 manuscript/manuscript.tex
Preparing for first public release
Revisions in response to reviews

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{2021-01-09}

@@

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase). However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here the performance of two multivariate extensions of the T-test are compared: Hotelling's \(T^2\) and a variant called \(T^2_{circ}\). A novel test of the assumptions of \(T^2_{circ}\) is developed, based on the condition index of the data (the square root of the ratio of eigenvalues of a bounding ellipse), and a heuristic for excluding outliers using the Mahalanobis distance is proposed. The \(T^2_{circ}\) statistic is then extended to multi-level designs, resulting in a new statistical test termed \(ANOVA^2_{circ}\). This has identical assumptions to \(T^2_{circ}\), and is shown to be more sensitive than MANOVA when these assumptions are met. The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run. Implementations of these novel tools are provided as an \emph{R} package, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

@@

A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal. The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-defined index of the brain's response (see Figure \ref{fig:fourierexplain}a,b). Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \ref{fig:fourierexplain}c), or equivalently as complex numbers with real and imaginary components (Figure \ref{fig:fourierexplain}d). In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data only. However an alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary components). Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods.

@@

where \(x_j\) denotes the \(j\)th observation of the dependent variables, and all other terms retain their previous meanings. Notice that no covariance term is present in equation \eqref{eq:t2c}, because of the independence assumption. This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as will be demonstrated below). Conceptually, this statistic takes the vector difference between the centroid (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the mean length of the residual vector lines joining each data point to the centroid. Two-sample and repeated measures versions of the \(T^2_{circ}\) statistic are also possible.

@@

\caption{Demonstration of skew in absolute Fourier amplitudes for signals of different strengths. Signal strength is quantified as Cohen's d, defined as the ratio of the mean to the standard deviation of the sample. The upper row shows samples of 50 grey points, and the population mean (coloured points). The lower row shows kernel density functions generated from 100,000 amplitude values. Note that the mean phase of the signal is irrelevant for these simulations, and is shown in the positive x-direction for consistency.}\label{fig:amphists}
\end{figure}

The upper row of Figure \ref{fig:amphists} shows scatterplots of simulated Fourier components, expressed using real (\emph{x}) and imaginary (\emph{y}) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \ref{fig:amphists} shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's \emph{d} \textgreater{} 2; Cohen's \emph{d} is the mean difference scaled by the standard deviation, see Cohen (1988)). One consequence of this is that T-tests will potentially have an inflated Type I error (false positive) rate for many signals encountered empirically, especially if used to make pointwise comparisons to an amplitude of 0.

Typical solutions, such as log-transforming the data, are unlikely to be equally applicable to all conditions. For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal. Applying a transform to both conditions is therefore problematic. Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents. Instead, the bivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.

@@

Using the \(T^2\) or \(T^2_{circ}\) statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values. Victor \& Mast (1991) report simulations showing situations where \(T^2_{circ}\) has greater power than \(T^2\). This involved generating random data sets of different sample sizes, and different signal strengths, and comparing the number of such tests where each statistic was significant. Their simulations show the largest advantage for \(T^2_{circ}\) for effect sizes around \emph{d} = 1 (where the mean signal strength is equal to the standard deviation of the data). The advantage appeared to be stronger for smaller sample sizes.

Here these simulations are replicated and extended (see Figure \ref{fig:powerfig}), and it is shown that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f). However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (when \emph{d} \textgreater{} 2 and N \textgreater{} 8). These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use should be restricted to small sample studies.

@@

Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on satisfying the \(T^2_{circ}\) test's more stringent assumptions. The two variables must be independent (i.e.~uncorrelated), and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \ref{fig:falsealarms} shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (horizontal dashed lines). However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the variances of the two dependent variables more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2. In contrast, the \(T^2\) statistic, which explicitly takes account of the covariance matrix (see equation \eqref{eq:t2eq}) shows no increase (black curves).

@@

One possible remedy to control the Type I error rate would be to adjust either the alpha level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated). However, this will reduce the statistical power of the \(T^2_{circ}\) test, and its advantage over \(T^2\) is relatively marginal in most situations to begin with (see Figure \ref{fig:powerfig}). What is required is a method to objectively assess whether the assumptions of \(T^2_{circ}\) hold; this is developed in the following section.

@@

\begin{equation}
\label{eq:edelman1}
pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)},
\end{equation}

Attempts to validate this by simulation suggest that for small sample sizes (\emph{N} \textless{} 10) a closer approximation is given by:

@@

Figure \ref{fig:distcomparison}a shows example distributions derived from both expressions (black and red curves), and simulations from 100,000 random data sets with \emph{N} = 4 (blue shading). The vertical lines show the critical (95\%) threshold for the analytic and simulated results. A ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance, because it has a condition index larger than expected by chance (assuming \(\alpha\) = 0.05). Ratios below the threshold imply that the eigenvalues can be considered equal (in a statistical sense). Figure \ref{fig:distcomparison}b shows how these critical thresholds change as a function of the number of observations (N), and it appears that the modified expression (red) most closely approximates the simulation results (blue).

@@

If a data set produces a significant result using the condition index test, this could be due to the presence of one or more outliers. The Mahalanobis distance (Mahalanobis, 1936) is a useful metric for identifying such multivariate outliers so that they can be excluded. It calculates the Euclidean distance between each data point and the centroid, and scales it by the variance in the direction of the vector that joins the two points. This means that any correlations in the data set are taken into account when calculating the distance metric, \emph{D}.

The effectiveness of this approach to outlier exclusion can be assessed by simulation using the condition index test. Figure \ref{fig:outlierplot} shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves). In all cases, the functions depart from the Type I error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the outlier's Mahalanobis distance exceeds a value around 3. This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R} (or \emph{mahal} function in \emph{Matlab}), return \(D^2\), which can be converted to \emph{D} by taking the square root). Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the \(T^2_{circ}\) test.

@@

Many studies involve more than two experimental conditions that need to be compared. Again, issues with familywise error will quickly become problematic if multiple pairwise \(T^2\) or \(T^2_{circ}\) statistics are calculated. One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's \(T^2\), but permits independent variables with more than two levels, as well as factorial designs. However, if the assumptions of \(T^2_{circ}\) hold for a data set, it might alternatively be possible to extend the logic of the \(T^2_{circ}\) test (Victor and Mast, 1991) to the more general case, and obtain a sensitivity benefit similar to that shown in Figure \ref{fig:powerfig}.

The F-statistic for a one-way ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance. For multivariate data, the change in group mean would be calculated using the vector distances between the complex means, and the residuals are the vector distances between each data point and its corresponding group mean. For an independent one-way design with \(k\) groups (or conditions) and \(N\) observations per group, the F-distribution will have \(2(k-1)\) and \(2((Nk)-k)\) degrees of freedom. The difference between traditional univariate ANOVA is the factor of two scaling, which accounts for the additional degree of freedom for each complex-valued number. A \emph{p}-value can then be derived from the F-distribution in the usual way.

@@

Following a significant \({ANOVA}^2_{circ}\) test, one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction). Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary; see Figure \ref{fig:fourierexplain}d). A repeated measures version of \({ANOVA}^2_{circ}\) can also be implemented following the same logic. The project repository contains a function (\emph{anovacirc.test} in \emph{R}, \emph{anovacirc\_test} in \emph{Matlab}) to run both of these tests for one-way designs. In principle factorial versions might also be derived.

@@

The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test. Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if they are not. Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Many MANOVA implementations cannot deal correctly with random factors (repeated measures), particularly in complex factorial designs. However the \emph{multRM} function in the \href{https://CRAN.R-project.org/package=MANOVA.RM}{\emph{MANOVA.RM}} package (Friedrich et al., 2019) is able to handle such designs appropriately using a bootstrapping approach.

@@

Figure \ref{fig:mousedata}a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz. There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality. Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (\emph{t} = 0.84, \emph{df} = 5, \emph{p} = 0.44). However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \ref{fig:mousedata}b). The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted. This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of \emph{D} = 2.14. This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al. (2019) went on to explore interactions between these two signals.

@@

Vilidaite et al. (2018) measured visual responses to flickering grating stimuli in a large sample of 100 adults. Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker). Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast. Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \ref{fig:humanSSVEP}. For the main analysis, responses were taken from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), and averaged across repetition for each participant. Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}. Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.

@@

\caption{Summary of human SSVEP data. Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes). Electrodes marked in red indicate cluster-corrected significance.  Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point). Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.}\label{fig:humanSSVEP}
\end{figure}

@@

It is worth stating explicitly that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations. This is the case for most paradigms in which the nervous system is driven by a periodic stimulus. However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger, 1929; Buzsáki and Draguhn, 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g. Canolty and Knight, 2010). When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al., 2021). This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.

@@

Even when statistics are conducted using both the real and imaginary Fourier components, it is still typical to visualise the mean amplitudes. Several approaches to calculating appropriate error bars have been proposed. For example, Pei et al. (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude. This approach is somewhat demanding, though a function is provided to calculate error bars using this method (available through the \emph{amperrors} function). However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes. This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean. The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled mean amplitudes. Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68\% and 95\% are typical, with the 68\% interval corresponding to the standard error).

@@

>>>>

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{2021-03-28}

@@

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase), either to test for the presence of a signal, or to compare signals across different conditions. However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here the performance of two multivariate extensions of the T-test are compared: Hotelling's \(T^2\) and a variant called \(T^2_{circ}\). A novel test of the assumptions of \(T^2_{circ}\) is developed, based on the condition index of the data (the square root of the ratio of eigenvalues of a bounding ellipse), and a heuristic for excluding outliers using the Mahalanobis distance is proposed. The \(T^2_{circ}\) statistic is then extended to multi-level designs, resulting in a new statistical test termed \(ANOVA^2_{circ}\). This has identical assumptions to \(T^2_{circ}\), and is shown to be more sensitive than MANOVA when these assumptions are met. The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run. Implementations of these novel tools are provided as an \emph{R} package and a \emph{Matlab} toolbox, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

@@

A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal. The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response (see Figure \ref{fig:fourierexplain}a,b). Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \ref{fig:fourierexplain}c), or equivalently as complex numbers with real and imaginary components (Figure \ref{fig:fourierexplain}d). In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data (or equivalently power or signal-to-noise ratios) only (e.g. Liu-Shuang et al., 2014; Afsari et al., 2014; Hou et al., 2020; Itthipuripat et al., 2014; McFadden et al., 2014; Smith et al., 2017; Vanegas et al., 2015) (note this is a non-exhaustive selection of recent studies). Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if signals differ between two (or more) experimental conditions. However, there are several objections to using univariate statistics to answer these questions, as will be demonstrated below. An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components). Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods for typical applications (detecting the presence of a signal, or comparing multiple signals).

@@

where \(x_j\) denotes the \(j\)th observation of the dependent variables, and all other terms retain their previous meanings. Notice that no covariance term is present in equation \eqref{eq:t2c}, because of the independence assumption. This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as will be demonstrated below). Conceptually, this statistic takes the vector difference between the bivariate sample mean (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the mean length of the residual vector lines joining each data point to the sample mean. Two-sample and repeated measures versions of the \(T^2_{circ}\) statistic are also possible.

@@

\caption{Demonstration of skew in absolute Fourier amplitudes for signals of different strengths. Signal strength is quantified as Cohen's d, defined as the ratio of the sample mean to the sample standard deviation. The upper row shows samples of 50 grey points, and the population mean (coloured points). The lower row shows kernel density functions generated from 100,000 amplitude values. Note that the mean phase of the signal is irrelevant for these simulations, and is shown in the positive x-direction for consistency.}\label{fig:amphists}
\end{figure}

The upper row of Figure \ref{fig:amphists} shows scatterplots of simulated Fourier components, expressed using real (\emph{x}) and imaginary (\emph{y}) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \ref{fig:amphists} shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's \emph{d} \textgreater{} 2; Cohen's \emph{d} is the difference of sample means scaled by the standard deviation, see Cohen (1988)). One consequence of this is that T-tests will potentially have an inflated Type I error (false positive) rate for many signals encountered empirically, especially if used to make pointwise comparisons to an amplitude of 0.

Typical solutions for dealing with skew, such as log-transforming the data, are unlikely to be equally applicable to all conditions. For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal. Applying a transform to both conditions is therefore problematic. Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents. Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.

@@

Using the \(T^2\) or \(T^2_{circ}\) statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values. Victor \& Mast (1991) report simulations showing situations where \(T^2_{circ}\) has greater power than \(T^2\), which is expected on theoretical grounds because the additional degrees of freedom in the \(T^2_{circ}\) test make it more efficient. Their simulations involved generating random data sets of different sample sizes, and different signal strengths (sample means), and comparing the number of such tests where each statistic was significant. The simulations showed the largest advantage for \(T^2_{circ}\) for effect sizes around \emph{d} = 1 (where the sample mean is equal to the sample standard deviation). The advantage appeared to be stronger for smaller sample sizes.

Here these simulations are replicated and extended (see Figure \ref{fig:powerfig}), and it is shown that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f). However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (when \emph{d} \textgreater{} 2 and N \textgreater{} 8). These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use can be restricted to small sample studies.

@@

Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on satisfying the \(T^2_{circ}\) test's more stringent assumptions. The two variables must be independent (i.e.~uncorrelated), and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \ref{fig:falsealarms} shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (horizontal dashed lines). However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the sample variances of the two dependent variables grow more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2. In contrast, the \(T^2\) statistic, which explicitly takes account of the covariance matrix (see equation \eqref{eq:t2eq}) shows no increase (black curves).

@@

One possible remedy to control the Type I error rate would be to adjust either the \(\alpha\) level or the degrees of freedom (as is done in repeated measures ANOVA when sphericity assumptions are violated). However, this will reduce the statistical power of the \(T^2_{circ}\) test, and its advantage over \(T^2\) is relatively marginal in most situations to begin with (see Figure \ref{fig:powerfig}). What is required is a method to objectively assess whether the assumptions of \(T^2_{circ}\) hold; this is developed in the following section.

@@

\begin{equation}
\label{eq:edelman1}
pdf = (N-1)2^{N-1}\frac{x^2 - 1}{(x^2 + 1)^N}x^{(N-2)}.
\end{equation}

Attempts to validate this by simulation (100,000 random data sets) found that for small sample sizes (\emph{N} \textless{} 10), there is a mismatch between the equation's predictions and the simulations (the black curve does not match the blue shading in Figure \ref{fig:distcomparison}a). Instead, a closer approximation to the simulations is given by:

@@

The red curve in Figure \ref{fig:distcomparison}a shows the predictions of this modified equation, which are much closer to the simulation results (blue shading). The vertical lines show the critical (95\%) threshold for the analytic and simulated results. A ratio lying beyond this threshold can be considered to violate the assumption of either independence or equal variance, because it has a condition index larger than expected by chance (assuming \(\alpha\) = 0.05). Ratios below the threshold imply that the eigenvalues can be considered equal (in a statistical sense). Figure \ref{fig:distcomparison}b shows how these critical thresholds change as a function of the number of observations (N), and it appears that the modified expression (red) most closely approximates the simulation results (blue).

@@

If a data set produces a significant result using the condition index test, this could be due to the presence of one or more outliers. The Mahalanobis distance (Mahalanobis, 1936) is a useful metric for identifying such multivariate outliers so that they can be excluded. It calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points. This means that any correlations in the data set are taken into account when calculating the distance metric, \emph{D}.

The effectiveness of this approach to outlier exclusion can be assessed by simulation using the condition index test. Figure \ref{fig:outlierplot} shows the proportion of significant condition index tests as a function of the Mahalanobis distance of a single outlier, for a range of sample sizes (curves). In all cases, the functions depart from the Type I error rate (\(\alpha\) = 0.05; horizontal dashed line in Figure \ref{fig:outlierplot}) when the outlier's Mahalanobis distance exceeds a value around 3. This seems a reasonable heuristic for outlier exclusion, and is the multivariate equivalent of excluding data points more than 3 standard deviations from the sample mean (note that many implementations of the Mahalanobis distance statistic, such as the core \emph{mahalanobis} function in \emph{R}, or \emph{mahal} function in \emph{Matlab}, return \(D^2\), which can be converted to \emph{D} by taking the square root). Following this heuristic should reduce the likelihood that outliers will invalidate the assumptions of the \(T^2_{circ}\) test.

@@

Many studies involve more than two experimental conditions that need to be compared. Again, issues with familywise error will quickly become problematic if multiple pairwise \(T^2\) or \(T^2_{circ}\) statistics are calculated. One possibility is to conduct a MANOVA, which takes covariances between dependent variables into account in much the same way as Hotelling's \(T^2\), but permits independent variables with more than two levels, as well as factorial designs. However, if the assumptions of \(T^2_{circ}\) hold for a data set, it should alternatively be possible to extend the logic of the \(T^2_{circ}\) test (Victor and Mast, 1991) to the more general multiple group case, and obtain a sensitivity benefit similar to that shown in Figure \ref{fig:powerfig}.

The F-statistic for a one-way independent ANOVA is calculated by taking the ratio between the variance explained by the modelled group means, and the residual unexplained variance:

\begin{equation}
\label{eq:Fratio}
F = \frac{MS_M}{MS_R},
\end{equation}

where \(MS_M\) is the mean squares of the linear model, and \(MS_R\) is the mean squares of the residuals. For bivariate data, the linear model is defined as the change in group (condition) means relative to the grand mean, calculated using the vector distances between points:

\begin{equation}
\label{eq:MSM}
MS_M = \frac{\Sigma N_k (\bar{x}_k - \bar{x}_{grand})^2}{df_M},
\end{equation}

where \(N_k\) is the number of observations in group \emph{k}, \(\bar{x}_k\) is the bivariate sample mean of group \emph{k}, \(\bar{x}_{grand}\) is the bivariate mean of the entire sample (including all groups), \(\Sigma\) denotes summation across groups, and \(df_M\) is the degrees of freedom for the model. The model degrees of freedom for \emph{k} groups is \(df_M = 2(k-1)\), with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.

The residuals are the vector distances between each data point and its corresponding group mean. The denominator component for the F-ratio equation is therefore defined as:

\begin{equation}
\label{eq:MSR}
MS_R = \frac{\Sigma (x_{i,k} - \bar{x}_k)^2}{df_R},
\end{equation}

where \(x_{i,k}\) is the \(i\)th data point from group \emph{k}, and other terms are as described previously. The degrees of freedom for a balanced design are calculated as \(df_R = 2((Nk)-k)\), where \emph{N} is the number of observations per group, and \emph{k} is the number of groups. A \emph{p}-value can then be determined by comparing the F-ratio calculated from equation \eqref{eq:Fratio} to an F-distribution with \(df_M\) and \(df_R\) degrees of freedom.

@@

Following a significant \({ANOVA}^2_{circ}\) test, one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction). Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes of the two measures depend on stimulus phase (which is arbitrary; see Figure \ref{fig:fourierexplain}d). A repeated measures version of \({ANOVA}^2_{circ}\) can also be implemented following the same logic. The project repository contains a function (\emph{anovacirc.test} in \emph{R}, \emph{anovacirc\_test} in \emph{Matlab}) to run both of these tests for one-way designs. In principle factorial versions might also be derived.

@@

The flowchart in Figure \ref{fig:flowchart} illustrates a proposed decision structure for the analysis of periodic data, once any outliers have been removed. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test. Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the condition index test is non-significant, and the \(T^2\) statistic otherwise. Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the condition index test is non-significant, or a MANOVA otherwise. Many MANOVA implementations cannot deal correctly with random factors (repeated measures), particularly in complex factorial designs. However the \emph{multRM} function in the \href{https://CRAN.R-project.org/package=MANOVA.RM}{\emph{MANOVA.RM}} package (Friedrich et al., 2019) is able to handle such designs appropriately using a bootstrapping approach.

@@

Figure \ref{fig:mousedata}a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz. There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality. Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (\emph{t} = 0.84, \emph{df} = 5, \emph{p} = 0.44). However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \ref{fig:mousedata}b). The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted. This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of \emph{D} = 2.14. This demonstrates that sound and (optogenetic) light entrain neural responses with different latencies (lags). The original study by Hwang et al. (2019) went on to explore interactions between these two signals.

@@

Vilidaite et al. (2018) measured visual responses to flickering grating stimuli in a large sample of 100 adults. Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker). Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast. Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent \(\alpha\) levels given the high power of this data set), as indicated by the red electrodes in the upper row of Figure \ref{fig:humanSSVEP}. For the main analysis, responses were taken from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), and averaged across repetition for each participant. Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}. Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.

@@

\caption{Summary of human SSVEP data. Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes). Electrodes marked in red indicate cluster-corrected significance.  Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (marked  by the black points in the scalp plots). Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the bivariate sample means.}\label{fig:humanSSVEP}
\end{figure}

@@

It is worth stating explicitly that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations. This is the case for most paradigms in which the nervous system is driven by a periodic stimulus, and sometimes also when phase-locked oscillatory responses are induced by a brief stimulus presentation (e.g.~in time-frequency analysis of event-related potentials, see Delorme and Makeig, 2004; Pfurtscheller and Aranibar, 1979). However, the methods are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger, 1929; Buzsáki and Draguhn, 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g. Canolty and Knight, 2010). When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al., 2021). This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.

@@

Even when statistics are conducted using both the real and imaginary Fourier components, it is still typical to visualise the mean amplitudes. Several approaches to calculating appropriate error bars have been proposed. For example, Pei et al. (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude. This approach is somewhat computationally demanding, though a function is provided to calculate error bars using this method (available through the \emph{amperrors} function). However, as an alternative, bootstrap resampling offers a powerful and general method for calculating confidence intervals on amplitudes. This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean. The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled mean amplitudes. Upper and lower confidence intervals on the amplitude can then be taken at appropriate quantiles of this population (68\% or 95\% are typical, with the 68\% interval corresponding to the standard error).

@@

\leavevmode\hypertarget{ref-Delorme2004}{}%
Delorme A, Makeig S. 2004. EEGLAB: An open source toolbox for analysis of single-trial EEG dynamics including independent component analysis. \emph{J Neurosci Methods} \textbf{134}:9--21. doi:\href{https://doi.org/10.1016/j.jneumeth.2003.10.009}{10.1016/j.jneumeth.2003.10.009}
